Rao-Blackwellised Two-filter algorithm for Conditionally Linear and Gaussian models

### 2016-11-01 TexArxiv/clgm.tex
minor modifictations in the numerical experiments

\begin{abstract}
Only a few methods have been proposed to reduce Monte Carlo variance of smoothers with Rao Blackwellisation steps for conditionally linear and Gaussian models. The most widely used Rao Blackwellised Sequential Monte Carlo smoothers introduced in \cite{kim:1994,lindsten:bunch:sarkka:schon:godsill:2015,sarkka:bunch:godsill:2012,lindsten:bunch:godsill:schon:2013} are based on the Forward Backward decomposition of the marginal smoothing distribution. In \cite{kim:1994}, a structural approximation of the model is used to simplify the backward pass while \cite{lindsten:bunch:sarkka:schon:godsill:2015,sarkka:bunch:godsill:2012,lindsten:bunch:godsill:schon:2013} use Rao Blackwellisation steps without any approximation in both forward and backward passes. In this paper, these algorithms are compared to an extension of the Two-filter based algorithm of \cite{briers:doucet:maskell:2010} with additional explicit Rao Blackwellised steps to improve its performance. The procedures are applied to commodity markets which are described using a two factor model based on the spot price and a convenience yield. Parameters are estimated using  an Expectation Maximization and the performance of the proposed procedure is assessed with Chicago Mercantile Exchange crude oil data.

@@

%Assuming that the switching rate between regimes is small compared to the inverse of the discretization period, the discretized version of the spot price and convenience yield $Z_k= (\ln S_k, \delta_k)$ (the sampling period is taken to be 1) is modeled as a 
State space models are bivariate stochastic processes $\{(Y_k,Z_k)\}_{k\ge 1}$ where the state sequence $\{Z_k\}_{k\ge 1}$ is a Markov chain which is only partially observed through the sequence $\{Y_k\}_{k\ge 1}$. Conditionally on the state sequence $\{Z_k\}_{k\ge 1}$ the observations are independent and for all $\ell\ge 1$ the conditional distribution of $Y_{\ell}$ given $\{Z_k\}_{k\ge 1}$ depends on $Z_{\ell}$ only. These models are used in a large variety of disciplines such as econometrics, target tracking, biology, signal processing, see \cite{delmoral:2013} and the references therein. In general state space models, bayesian filtering and smoothing problems, i.e. computation of the posterior distributions of a sequence of states $(Z_{k},\ldots,Z_{p})$ for $1\le k\le p\le \ell$ given observations $(Y_{1},\ldots,Y_{\ell})$, are challenging tasks required for instance  to perform maximum likelihood inference. Filtering refers to the estimation of the distributions of $Z_k$ given the observations $(Y_1,\ldots,Y_k)$ up to time k, while smoothing is the estimation of the distribution of states $(Z_{k},\ldots,Z_{p})$ given observations $(Y_{1},\ldots,Y_{\ell})$ with $1\le k\le p<\ell$. These distributions are in particular crucial to infer a sequence of states given the observations in real world online learning applications, see for instance \cite{thrun:burgard:fox:2005} for Simultaneous Localization and Mapping problems, \cite{barshalom:li:kirubarjan:2001} for target tracking problems or \cite{sarkka:2013,douc:moulines:stoffer:2013} for other applications in engineering, telecommunications... In the case where the state and observation models are linear and Gaussian, the filtering and smoothing problems may be solved optimally using the Kalman filter or the Rauch-Tung-Striebel smoother of \cite{kalman:1960} and \cite{rauch:striebel:tung:1965}. Many extensions of these algorithms based on local linearization methods have been proposed for nonlinear dynamical systems such as extended Kalman filters, iterated Kalman filters or Gaussian sum filters...

@@

An important feature of these models is that, conditionally on a trajectory $(a_1,\ldots,a_n)$ of indicator variables, both the state equation and the observation equation are linear and Gaussian. To exploit this specific structure using Rao Blackwellisation techniques, an alternative approach to linearization techniques to solve filtering and smoothing problems is to use Sequential Monte Carlo (SMC) methods. These methods based on sequential importance sampling and resampling techniques introduced in \cite{handschin:mayne:1969} and \cite{rubin:1987} allow to approximate any sequence of conditional distributions of the hidden states given the observations, see \cite{kitagawa:1987,gordon:salmond:smith:1993}. By using the tractable substructure of CLGM \cite{chen:liu:2000,doucet:godsill:andrieu:2000,doucet:gordon:krishnamurthy:2001,schon:gustafsson:nordlund:2005} introduced several SMC methods combined with Kalman filters to address the filtering problem. These algorithms sample only trajectories of indicator variables associated with importance weights, the corresponding state variables $(Z_1,\ldots,Z_k)$ are marginalized and estimated with a Kalman filter, which is the optimal filter in this case. However, many maximum likelihood based estimation procedures such as  the Expectation Maximization  (EM)  algorithm introduced in \cite{dempster:laird:rubin:1977} require to approximate some marginal smoothing distributions of the states and indicator variables. Several algorithms have been developed to solve the smoothing problem using SMC methods: for instance the Forward Filtering Backward Smoothing algorithm (FFBS), the Forward Filtering Backward Simulation algorithm (FFBSi) and the PArticle-based, Rapid Incremental Smoother (PaRIS)  developed respectively in \cite{kitagawa:1996,huerzeler:kunsch:1998,doucet:godsill:andrieu:2000},  \cite{godsill:doucet:west:2004} and \cite{olsson:westerborn:2015}. However, only a few methods have been proposed to reduce Monte Carlo variance of smoothers with Rao Blackwellisation steps. Instead of using a structural approximation of the model such as in \cite{kim:1994}, \cite{lindsten:bunch:sarkka:schon:godsill:2015,sarkka:bunch:godsill:2012,lindsten:bunch:godsill:schon:2013} proposed a Rao Blackwellised smoother based on the forward-backward decomposition of the FFBS algorithm with Rao Blackwellisation steps both in the forward and backward time directions. In the forward pass of these algorithms, a Rao Blackwellised particle filter such as \cite{chen:liu:2000} is used to approximate the filtering distributions and the corresponding particles, weights and associated trajectories of means and variances of the Kalman filters are stored. The backward pass samples trajectories of indicator variables backward in time using the particles and weights of the forward pass. \cite{briers:doucet:maskell:2010} proposed another Rao Blackwellised smoother which extends the Bryson-Frazier smoother for Gaussian linear state space models using the generalized two-filter formula with Rao Blackwellisation steps for the forward and the backward filters.  This two-filter approach combines a forward and a backward filters which are approximated numerically using SMC and Kalman filtering techniques. In \cite{briers:doucet:maskell:2010}, for all $1\le i\le n$, the sampled forward and backward sequences are merged to approximate the posterior distribution of $(a_i,z_i)$. In this paper, we propose to approximate the marginal smoothing distribution of $(a_i,z_i)$ by merging the sampled forward and backward trajectories at times $i-1$ and $i+1$ and integrating out all possible paths between time $i-1$ and time $i$ and between time $i$ and time $i+1$ instead of sampling random variables. This explicit marginalization provides an additional Rao Blackwellisation step and produce more robust approximations.

The paper is organized as follows. The proposed Monte Carlo algorithm to estimate the marginal smoothing distributions of the hidden states is presented in Section~\ref{sec:RaoBlackwell} along with the other widely used Rao Blackwellised smoothing algorithms. The performance of all these methods is illustrated in Section~\ref{sec:numerical:experiments} with simulated data. Then, the parameters of the proposed model are estimated using a Monte Carlo EM algorithm where the required conditional expectations are replaced by their estimates based on the proposed SMC method.  
The performance of this algorithm is assessed using simulated data and CME crude oil (WTI) data. Intermediate computations and explicit approximations provided by the algorithm are given in the Appendix.

\section{Rao Blackwellised smoothing algorithms}
\label{sec:RaoBlackwell}
This section details the Sequential Monte Carlo algorithms which can be used to approximate the conditional distribution of the states $(a_1,\ldots,a_n)$ or the marginal distributions of $(a_i,z_{i})$ given the observations $(Y_1,\ldots,Y_n)$. For all $m\,\times\,m$ positive-definite matrix $A$ and  all $z\in\mathbb{R}^{m}$ define
\[
\normMat{A}{z}^2 \eqdef z'A^{-1}z\eqsp.
\]
Let $m_{}(a_{i},z_{i-1};z_i)$ be the probability density of the conditional distribution of $Z_i$ given $(a_{i},Z_{i-1})$ and $g_{}(a_{i},z_{i};y_i)$ be the probability density of the conditional distribution of $Y_i$ given $(a_{i},Z_{i})$:
\begin{align}
\label{eq:definition-m}
m_{}(a_{i},z_{i-1};z_i) & \eqdef \left|2\pi\barH_{a_i}\right|^{-1/2}\exp\left\{-\frac{1}{2}\normMat{\barH_{a_i}}{z_i -d_{a_{i}} -T_{a_{i}}z_{i-1}}^2\right\}\eqsp,\\
\label{eq:definition-g}
g_{}(a_{i},z_{i};y_i) & \eqdef \left|2\pi\barG_{a_i}\right|^{-1/2}\exp\left\{-\frac{1}{2}\normMat{\barG_{a_i}}{y_i - c_{a_i} - B_{a_i}z_{i}}^2\right\}\eqsp,
\end{align}
where
\[
\barG_j \eqdef G_jG'_j\;,\; \barH_{j} \eqdef H_jH'_j\eqsp.
\]
%Maximum likelihood estimation of the parameters using the Expectation Maximization algorithm or gradient ascent algorithms requires to compute the posterior distributions of the states $a_{i-1:i},z_{i-1:i}$  given the observations $Y_{1:n}$, written $\phi^n_{i}(a_{i-1:i},z_{i-1:i})$, which cannot be computed in closed-form (the dependency of $\phi^n_{i}$ on $Y_{1:n}$ is dropped for better clarity). SMC methods provide a class of importance sampling and resampling procedures to approximate $\phi^n_{i}(a_{i-1:i},z_{i-1:i})$. 
All the algorithms considered in this paper are based on forward-backward or two-filter decompositions of the smoothing distribution and share the same forward filter detailed in Section~\ref{sec:forward}.

\subsection{Forward filter}
\label{sec:forward}
The SMC approximation $p^N_{}(a_{1:i},z_{i}|y_{1:i})$ of $p_{}(a_{1:i},z_{i}|y_{1:i})$ may be obtained using the standard Rao Blackwellised  algorithm given in \cite{chen:liu:2000}. The procedure produces a sequence of trajectories $(a^k_{1:i})_{1\le k \le N}$ associated with importance weights $(\omega^k_{i})_{1\le k \le N}$ used to define the following approximation of $p_{}(a_{1:i},z_{i}|y_{1:i})$:
\[
p^N_{}(a_{1:i},z_{i}|y_{1:i}) = \sum_{k=1}^N\delta_{a^k_{1:i}}(a_{1:i})\,\omega^k_{i}\,p_{}(z_{i}|a^k_{1:i},y_{1:i})\eqsp.
\]
In this equation, the conditional distribution of the hidden state $z_{i}$ given the observations $y_{1:i}$ and a trajectory $a^k_{1:i}$ is a Gaussian distribution whose mean $\mu^k_{i}$ and variance $P^k_{i}$ may be obtained by running a Kalman filter. 
\paragraph{Initialization}
At time $i=1$, $(a^k_1)_{1\le k \le N}$ are sampled independently in $\{1,\ldots,J\}$ with probabilities proportional to $(\pi_j)_{1\le j\le J}$. Then, $\mu_1^k$ and $P_1^k$ are computed using a Kalman filter:
\begin{align*}
K^k_{1} &=\Sigma_1B'_{a_1^k}\left(B_{a_1^k}\Sigma_1B'_{a_1^k} + \barG_{a_1^k}\right)^{-1}\eqsp,\\
\mu^k_{1} &= \mu_{1} + K^k_1\left(Y_1 - c_{a_{1}^k} - B_{a_1^k}\mu_{1}\right)\eqsp,\\
P^k_{1} &=\left(I-K^k_{1}B_{a_1^k}\right)\Sigma_1\eqsp.
\end{align*}
Particle $a_1^k$ is associated with the importance weight $\omega^k_1 \propto p(y_1|a_1^k)$. This weight is updated by noting that the conditional distribution of $Y_1$ given $a_1^k$ is a Gaussian distribution with mean $c_{a^k_1}+B_{a^k_1}\mu_1$ and variance $B_{a^k_1}\Sigma_1B'_{a^k_1} + \barG_{a^k_1}$. 
\paragraph{Iterations}
For all $2\le i \le n$, particle indices $(I^k_i)_{1\le k \le N}$ are sampled independently in $\{1,\ldots,N\}$ with probabilities proportional to  $(\omega^k_{i-1})_{1\le k \le N}$.  Labels trajectories at time $i-1$ are defined by:
\[
a_{1:i-1}^k = (a^{\mathsf{A}^k_1}_1,\ldots,a^{\mathsf{A}^k_{i-1}}_{i-1})\eqsp,
\]
where
\[
\mathsf{A}^k_{i-1} = I^k_i \quad\mbox{and for } 1\le j<i\quad \mathsf{A}^k_{j} = I^{\mathsf{A}^k_{j+1}}_{j+1}\eqsp.
\]
Similarly, means and variances trajectories are defined by: 
\[
\mu_{1:i-1}^k = (\mu^{\mathsf{A}^k_1}_1,\ldots,\mu^{\mathsf{A}^k_{i-1}}_{i-1})\quad\mbox{and}\quad P_{1:i-1}^k = (P^{\mathsf{A}^k_1}_1,\ldots,P^{\mathsf{A}^k_{i-1}}_{i-1})\eqsp.
\]
The new indicator variables $(a_i^k)_{1\le k \le N}$ are sampled independently in  $\{1,\ldots, J\}$ with probabilities proportional to $(\gamma_i^{j,k})_{1\le j\le J}$, where for all $j \in \{1,\ldots, J\}$:
\[
\gamma_i^{j,k} = p(y_i | a_i = j, a_{1:i-1}^k, y_{1:i-1}) Q(a_{i-1}^k,j)\eqsp.
\]
The conditional distribution of $Y_i$ given $a^k_{1:i-1}$, $a_i$ and $Y_{1:i-1}$ is a Gaussian distribution with mean $c_{a_i}+B_{a_i}\mu^k_{i|i-1}$ and variance $B_{a_i}P^k_{i|i-1}B'_{a_i} + \barG_{a_i}$ which means that $\gamma_i^{j,k}$ may be written:
\[
\gamma_i^{j,k} =  Q(a_{i-1}^k,j)|B_jP^k_{i|i-1}B'_j + \barG_j|^{-1/2}\exp\left\{-\frac{1}{2}\normMat{B_jP^k_{i|i-1}B'_j + \barG_j}{y_i-c_{j}-B_j\mu^k_{i|i-1}}^2\right\} \eqsp,
\]
where 
\begin{align*}
\mu^{k}_{i|i-1} &= d_{a_{i}^k} + T_{a_{i}^k}\mu^k_{i-1}\eqsp,\\
P^k_{i|i-1} &= T_{a_{i}^k}P^k_{i-1}T'_{a_{i}^k} + \barH_{a_{i}^k}\eqsp.
\end{align*}
$\mu_i^k$ and $P_i^k$ are updated according to:
\begin{align*}
K^k_{i} &=P^k_{i|i-1}B'_{a_{i}^k}\left(B_{a_{i}^k}P^k_{i|i-1}B'_{a_{i}^k} + \barG_{a_{i}^k}\right)^{-1}\eqsp,\\
\mu^k_{i} &= \mu^{k}_{i|i-1} + K^k_i\left(Y_i - c_{a_{i}^k} - B_{a_{i}^k}\mu^{k}_{i|i-1}\right)\eqsp,\\
P^k_{i} &=\left(I-K^k_{i}B_{a_{i}^k}\right)P^k_{i|i-1}\eqsp.
\end{align*}
The importance weights are given by $\omega_i^k \propto \sum_{j=1}^J \gamma_i^{j,k}$.

@@

\subsection{FFBS based algorithm of \cite{lindsten:bunch:sarkka:schon:godsill:2015}}
\cite{lindsten:bunch:sarkka:schon:godsill:2015} proposed a Rao-Blackwellised procedure to sample the indicator variables backward in time following the same steps as in the FFBS algorithm without using the approximation of \cite{kim:1994}. The algorithm relies on the decomposition given, for all $1\le i \le n-1$, by:

@@

\subsection{Rao Blackwellised Two-filter Smoother}
Following \cite{briers:doucet:maskell:2010}, this paper proposes a generalized two-filter algorithm which exploits the specific properties of the model defined by \eqref{eq:model:state} and \eqref{eq:model:obs}. Contrary to the previous methods, two-filter based smoothers are designed to compute approximations of marginal smoothing distributions (usually the posterior distribution of one or two consecutive  indicator variables given all the observations). 

@@

For $3\le i \le n-1$, our proposed algorithm relies on the explicit marginalization:

@@

\phi^n_{i}(a_{i},z_{i})  =  \sum_{a_{i-1}}\sum_{a_{i+1}} \int_{z_{i-1}} \int_{z_{i+1}} \psi^n_{i}(a_{i-1:i+1},z_{i-1:i+1})\rmd z_{i-2}\rmd z_{i+1}\eqsp.

@@

where $\psi^n_{i}(a_{i-1:i+1},z_{i-1:i+1})$ is the smoothing distribution of the hidden regimes and states between time indices $i-1$ and $i+1$.
It may be decomposed as follows:

@@

\clearpage
\newpage

>>>>

\begin{abstract}
This paper focuses on Sequential Monte Carlo based approximations of smoothing distributions in conditionally linear and Gaussian state spaces. Only a few methods have been proposed to reduce Monte Carlo variance of smoothers with Rao Blackwellisation steps for conditionally linear and Gaussian models. The most widely used Rao Blackwellised Sequential Monte Carlo smoothers introduced in \cite{kim:1994,lindsten:bunch:sarkka:schon:godsill:2015,sarkka:bunch:godsill:2012,lindsten:bunch:godsill:schon:2013} are based on the Forward Backward decomposition of the marginal smoothing distribution. In \cite{kim:1994}, a structural approximation of the model is used to simplify the backward pass while \cite{lindsten:bunch:sarkka:schon:godsill:2015,sarkka:bunch:godsill:2012,lindsten:bunch:godsill:schon:2013} use Rao Blackwellisation steps without any approximation in both forward and backward passes. In this paper, these algorithms are compared to an extension of the Two-filter based algorithm of \cite{briers:doucet:maskell:2010} with additional explicit Rao Blackwellised steps to improve its performance. The procedures are applied to commodity markets which are described using a two factor model based on the spot price and a convenience yield. Parameters are estimated using  an Expectation Maximization and the performance of the proposed procedure is assessed with Chicago Mercantile Exchange crude oil data.

@@

%Assuming that the switching rate between regimes is small compared to the inverse of the discretization period, the discretized version of the spot price and convenience yield $Z_k= (\ln S_k, \delta_k)$ (the sampling period is taken to be 1) is modeled as a 
State space models are bivariate stochastic processes $\{(Y_k,Z_k)\}_{k\ge 1}$ where the state sequence $\{Z_k\}_{k\ge 1}$ is a Markov chain which is only partially observed through the sequence $\{Y_k\}_{k\ge 1}$. Conditionally on the state sequence $\{Z_k\}_{k\ge 1}$ the observations are independent and for all $\ell\ge 1$ the conditional distribution of $Y_{\ell}$ given $\{Z_k\}_{k\ge 1}$ depends on $Z_{\ell}$ only. These models are used in a large variety of disciplines such as econometrics, target tracking, biology, signal processing, see \cite{delmoral:2013} and the references therein. In general state space models, bayesian filtering and smoothing problems, i.e. the computation of the posterior distributions of a sequence of states $(Z_{k},\ldots,Z_{p})$ for $1\le k\le p\le \ell$ given observations $(Y_{1},\ldots,Y_{\ell})$, are challenging tasks required for instance  to perform maximum likelihood inference. Filtering refers to the estimation of the distributions of $Z_k$ given the observations $(Y_1,\ldots,Y_k)$ up to time k, while smoothing is the estimation of the distribution of states $(Z_{k},\ldots,Z_{p})$ given observations $(Y_{1},\ldots,Y_{\ell})$ with $1\le k\le p<\ell$. These distributions are in particular crucial to infer a sequence of states given the observations in real world online learning applications, see for instance \cite{thrun:burgard:fox:2005} for Simultaneous Localization and Mapping problems, \cite{barshalom:li:kirubarjan:2001} for target tracking problems or \cite{sarkka:2013,douc:moulines:stoffer:2013} for other applications in engineering, telecommunications... In the case where the state and observation models are linear and Gaussian, the filtering and smoothing problems may be solved optimally using the Kalman filter or the Rauch-Tung-Striebel smoother of \cite{kalman:1960} and \cite{rauch:striebel:tung:1965}. Many extensions of these algorithms based on local linearization methods have been proposed for nonlinear dynamical systems such as extended Kalman filters, iterated Kalman filters or Gaussian sum filters...

@@

An important feature of these models is that, conditionally on a trajectory $(a_1,\ldots,a_n)$ of indicator variables, both the state equation and the observation equation are linear and Gaussian. To exploit this specific structure using Rao Blackwellisation techniques, an alternative approach to linearization techniques to solve filtering and smoothing problems is to use Sequential Monte Carlo (SMC) methods. These methods based on sequential importance sampling and resampling techniques introduced in \cite{handschin:mayne:1969} and \cite{rubin:1987} allow to approximate any sequence of conditional distributions of the hidden states given the observations, see \cite{kitagawa:1987,gordon:salmond:smith:1993}. By using the tractable substructure of CLGM \cite{chen:liu:2000,doucet:godsill:andrieu:2000,doucet:gordon:krishnamurthy:2001,schon:gustafsson:nordlund:2005} introduced several SMC methods combined with Kalman filters to address the filtering problem. These algorithms sample only trajectories of indicator variables associated with importance weights, the corresponding state variables $(Z_1,\ldots,Z_k)$ are marginalized and estimated with a Kalman filter, which is the optimal filter in this case. However, many maximum likelihood based estimation procedures such as  the Expectation Maximization  (EM)  algorithm introduced in \cite{dempster:laird:rubin:1977} require to approximate some marginal smoothing distributions of the states and indicator variables. Several algorithms have been developed to solve the smoothing problem using SMC methods: for instance the Forward Filtering Backward Smoothing algorithm (FFBS), the Forward Filtering Backward Simulation algorithm (FFBSi) and the PArticle-based, Rapid Incremental Smoother (PaRIS)  developed respectively in \cite{kitagawa:1996,huerzeler:kunsch:1998,doucet:godsill:andrieu:2000},  \cite{godsill:doucet:west:2004} and \cite{olsson:westerborn:2015}. However, only a few methods have been proposed to reduce Monte Carlo variance of smoothers with Rao Blackwellisation steps. Instead of using a structural approximation of the model such as in \cite{kim:1994}, \cite{fong:godsill:doucet:west:2002,lindsten:schon:2011} proposed forward-backward smoothers which require to sample the hidden linear states in the backward pass. \cite{lindsten:bunch:sarkka:schon:godsill:2015,sarkka:bunch:godsill:2012,lindsten:bunch:godsill:schon:2013} proposed a Rao Blackwellised smoother based on the forward-backward decomposition of the FFBS algorithm with Rao Blackwellisation steps both in the forward and backward time directions. In the forward pass of these algorithms, a Rao Blackwellised particle filter such as \cite{chen:liu:2000} is used to approximate the filtering distributions and the corresponding particles, weights and associated trajectories of means and variances of the Kalman filters are stored. The backward pass samples trajectories of indicator variables backward in time using the particles and weights of the forward pass. \cite{briers:doucet:maskell:2010} proposed another Rao Blackwellised smoother which extends the Bryson-Frazier smoother for Gaussian linear state space models using the generalized two-filter formula with Rao Blackwellisation steps for the forward and the backward filters.  This two-filter approach combines a forward and a backward filters which are approximated numerically using SMC and Kalman filtering techniques. In \cite{briers:doucet:maskell:2010}, for all $1\le i\le n$, the sampled forward and backward sequences are merged to approximate the posterior distribution of $(a_i,z_i)$. In this paper, we propose to approximate the marginal smoothing distribution of $(a_i,z_i)$ by merging the sampled forward and backward trajectories at times $i-1$ and $i+1$ and integrating out all possible paths between time $i-1$ and time $i$ and between time $i$ and time $i+1$ instead of sampling random variables. This explicit marginalization provides an additional Rao Blackwellisation step and produces more robust approximations.

The paper is organized as follows. An overview of the proposed Monte Carlo algorithms to estimate the smoothing distributions of the hidden states is presented in Section~\ref{sec:RaoBlackwell}. The performance of all these methods is illustrated in Section~\ref{sec:numerical:experiments} with simulated data. Then, a new model is proposed to explain regime switchings for commodity markets. The parameters of the proposed model are estimated using a Monte Carlo EM algorithm where the required conditional expectations are replaced by their estimates based on the proposed SMC method.  
The performance of the extended Rao-Blackwellised two-filter  algorithm is assessed using simulated data and CME crude oil (WTI) data. Intermediate computations and explicit approximations provided by the algorithm are given in the Appendix.

\section{Rao Blackwellised smoothing algorithms}
\label{sec:RaoBlackwell}
This section details the Sequential Monte Carlo algorithms which can be used to approximate the conditional distribution of the states $(a_1,\ldots,a_n)$ or the marginal distributions of $(a_i,z_{i})$ given the observations $(Y_1,\ldots,Y_n)$. For all $m\,\times\,m$ positive-definite matrix $A$ and  all $z\in\mathbb{R}^{m}$ define
\[
\normMat{A}{z}^2 \eqdef z'A^{-1}z\eqsp.
\]
Let $m_{}(a_{i},z_{i-1};z_i)$ be the probability density of the conditional distribution of $Z_i$ given $(a_{i},Z_{i-1})$ and $g_{}(a_{i},z_{i};y_i)$ be the probability density of the conditional distribution of $Y_i$ given $(a_{i},Z_{i})$:
\begin{align}
\label{eq:definition-m}
m_{}(a_{i},z_{i-1};z_i) & \eqdef \left|2\pi\barH_{a_i}\right|^{-1/2}\exp\left\{-\frac{1}{2}\normMat{\barH_{a_i}}{z_i -d_{a_{i}} -T_{a_{i}}z_{i-1}}^2\right\}\eqsp,\\
\label{eq:definition-g}
g_{}(a_{i},z_{i};y_i) & \eqdef \left|2\pi\barG_{a_i}\right|^{-1/2}\exp\left\{-\frac{1}{2}\normMat{\barG_{a_i}}{y_i - c_{a_i} - B_{a_i}z_{i}}^2\right\}\eqsp,
\end{align}
where
\[
\barG_j \eqdef G_jG'_j\;,\; \barH_{j} \eqdef H_jH'_j\eqsp.
\]
%Maximum likelihood estimation of the parameters using the Expectation Maximization algorithm or gradient ascent algorithms requires to compute the posterior distributions of the states $a_{i-1:i},z_{i-1:i}$  given the observations $Y_{1:n}$, written $\phi^n_{i}(a_{i-1:i},z_{i-1:i})$, which cannot be computed in closed-form (the dependency of $\phi^n_{i}$ on $Y_{1:n}$ is dropped for better clarity). SMC methods provide a class of importance sampling and resampling procedures to approximate $\phi^n_{i}(a_{i-1:i},z_{i-1:i})$. 
All the algorithms considered in this paper are based on forward-backward or two-filter decompositions of the smoothing distributions and share the same forward filter detailed in Section~\ref{sec:forward}.

\subsection{Forward filter}
\label{sec:forward}
The SMC approximation $p^N_{}(a_{1:i},z_{i}|y_{1:i})$ of $p_{}(a_{1:i},z_{i}|y_{1:i})$ may be obtained using the standard Rao Blackwellised  algorithm given in \cite{chen:liu:2000}. The procedure produces a sequence of trajectories $(a^k_{1:i})_{1\le k \le N}$ associated with importance weights $(\omega^k_{i})_{1\le k \le N}$ used to define the following approximation of $p_{}(a_{1:i},z_{i}|y_{1:i})$:
\[
p^N_{}(a_{1:i},z_{i}|y_{1:i}) = \sum_{k=1}^N\delta_{a^k_{1:i}}(a_{1:i})\,\omega^k_{i}\,p_{}(z_{i}|a^k_{1:i},y_{1:i})\eqsp.
\]
In this equation, the conditional distribution of the hidden state $z_{i}$ given the observations $y_{1:i}$ and a trajectory $a^k_{1:i}$ is a Gaussian distribution whose mean $\mu^k_{i}$ and variance $P^k_{i}$ may be obtained by running a Kalman filter as follows. 
\paragraph{Initialization}
At time $i=1$, $(a^k_1)_{1\le k \le N}$ are sampled independently in $\{1,\ldots,J\}$ with probabilities proportional to $(\pi_j)_{1\le j\le J}$. Then, $\mu_1^k$ and $P_1^k$ are computed using a Kalman filter:
\begin{align*}
K^k_{1} &=\Sigma_1B'_{a_1^k}\left(B_{a_1^k}\Sigma_1B'_{a_1^k} + \barG_{a_1^k}\right)^{-1}\eqsp,\\
\mu^k_{1} &= \mu_{1} + K^k_1\left(Y_1 - c_{a_{1}^k} - B_{a_1^k}\mu_{1}\right)\eqsp,\\
P^k_{1} &=\left(I-K^k_{1}B_{a_1^k}\right)\Sigma_1\eqsp.
\end{align*}
Particle $a_1^k$ is associated with the importance weight $\omega^k_1 \propto p(y_1|a_1^k)$. This weight is updated by noting that the conditional distribution of $Y_1$ given $a_1^k$ is a Gaussian distribution with mean $c_{a^k_1}+B_{a^k_1}\mu_1$ and variance $B_{a^k_1}\Sigma_1B'_{a^k_1} + \barG_{a^k_1}$. 
\paragraph{Iterations}
For all $2\le i \le n$, particle indices $(I^k_i)_{1\le k \le N}$ are sampled independently in $\{1,\ldots,N\}$ with probabilities proportional to  $(\omega^k_{i-1})_{1\le k \le N}$.  Labels trajectories at time $i-1$ are defined by:
\[
a_{1:i-1}^k = (a^{\mathsf{A}^k_1}_1,\ldots,a^{\mathsf{A}^k_{i-1}}_{i-1})\eqsp,
\]
where
\[
\mathsf{A}^k_{i-1} = I^k_i \quad\mbox{and for } 1\le j<i\quad \mathsf{A}^k_{j} = I^{\mathsf{A}^k_{j+1}}_{j+1}\eqsp.
\]
Similarly, means and variances trajectories are defined by: 
\[
\mu_{1:i-1}^k = (\mu^{\mathsf{A}^k_1}_1,\ldots,\mu^{\mathsf{A}^k_{i-1}}_{i-1})\quad\mbox{and}\quad P_{1:i-1}^k = (P^{\mathsf{A}^k_1}_1,\ldots,P^{\mathsf{A}^k_{i-1}}_{i-1})\eqsp.
\]
The new indicator variables $(a_i^k)_{1\le k \le N}$ are sampled independently in  $\{1,\ldots, J\}$ with probabilities proportional to $(\gamma_i^{j,k})_{1\le j\le J}$, where for all $j \in \{1,\ldots, J\}$:
\[
\gamma_i^{j,k} = p(y_i | a_i = j, a_{1:i-1}^k, y_{1:i-1}) Q(a_{i-1}^k,j)\eqsp.
\]
The conditional distribution of $Y_i$ given $a^k_{1:i-1}$, $a_i$ and $Y_{1:i-1}$ is a Gaussian distribution with mean $c_{a_i}+B_{a_i}\mu^k_{i|i-1}$ and variance $B_{a_i}P^k_{i|i-1}B'_{a_i} + \barG_{a_i}$ which means that $\gamma_i^{j,k}$ may be written:
\[
\gamma_i^{j,k} =  Q(a_{i-1}^k,j)|B_jP^k_{i|i-1}B'_j + \barG_j|^{-1/2}\exp\left\{-\frac{1}{2}\normMat{B_jP^k_{i|i-1}B'_j + \barG_j}{y_i-c_{j}-B_j\mu^k_{i|i-1}}^2\right\} \eqsp,
\]
where 
\begin{align*}
\mu^{k}_{i|i-1} &= d_{a_{i}^k} + T_{a_{i}^k}\mu^k_{i-1}\eqsp,\\
P^k_{i|i-1} &= T_{a_{i}^k}P^k_{i-1}T'_{a_{i}^k} + \barH_{a_{i}^k}\eqsp.
\end{align*}
$\mu_i^k$ and $P_i^k$ are updated according to:
\begin{align*}
K^k_{i} &=P^k_{i|i-1}B'_{a_{i}^k}\left(B_{a_{i}^k}P^k_{i|i-1}B'_{a_{i}^k} + \barG_{a_{i}^k}\right)^{-1}\eqsp,\\
\mu^k_{i} &= \mu^{k}_{i|i-1} + K^k_i\left(Y_i - c_{a_{i}^k} - B_{a_{i}^k}\mu^{k}_{i|i-1}\right)\eqsp,\\
P^k_{i} &=\left(I-K^k_{i}B_{a_{i}^k}\right)P^k_{i|i-1}\eqsp.
\end{align*}
The importance weights are given by $\omega_i^k \propto \sum_{j=1}^J \gamma_i^{j,k}$.

@@

\subsection{FFBS based algorithm of \cite{lindsten:bunch:sarkka:schon:godsill:2015,sarkka:bunch:godsill:2012,lindsten:bunch:godsill:schon:2013}}
\cite{lindsten:bunch:sarkka:schon:godsill:2015,sarkka:bunch:godsill:2012,lindsten:bunch:godsill:schon:2013} proposed a Rao-Blackwellised procedure to sample the indicator variables backward in time following the same steps as in the FFBS algorithm without using the approximation of \cite{kim:1994}. The algorithm relies on the decomposition given, for all $1\le i \le n-1$, by:

@@

\subsection{Rao Blackwellised Two-filter Smoother (RBTF)}
Following \cite{briers:doucet:maskell:2010}, this paper proposes a generalized two-filter algorithm which exploits the specific properties of the model defined by \eqref{eq:model:state} and \eqref{eq:model:obs}. Contrary to the previous methods, two-filter based smoothers are designed to compute approximations of marginal smoothing distributions (usually the posterior distribution of one or two consecutive  indicator variables given all the observations). \cite{briers:doucet:maskell:2010} introduced the following decomposition of the smoothing distributions for all $2\le i \le n$:
\label{eq:dens:bwd_fwd_doucet}
p(a_{i},z_{i}|y_{1:n})  \propto  p_{}(a_{i},z_{i}|y_{1:i-1})p(y_{i:n}|a_i,z_i)\eqsp.
The first term in the right hand side may be approximated using the forward filter, however, as the function $(a_{i},z_{i})\mapsto p_{}(y_{i:n}|a_{i},z_{i})$ is not a probability density function, it is not possible to use SMC samples to approximate it. The backward filter uses artificial densities to introduce a surrogate target density function which may be approximated recursively using SMC methods. Then, the forward and backward weighted samples are combined using \eqref{eq:dens:bwd_fwd_doucet} to approximate $p(a_{i},z_{i}|y_{1:n})$. %Following \cite{briers:doucet:maskell:2010}, for any probability densities $(\gamma^{}_{i})_{1\le i \le n}$, define the following joint probability densities:

@@

For $2\le i \le n-1$, our proposed algorithm relies on the explicit marginalization:

@@

p(a_{i},z_{i}|y_{1:n})  =  \sum_{a_{i-1}}\sum_{a_{i+1}} \int_{z_{i-1}} \int_{z_{i+1}} \psi^n_{i}(a_{i-1:i+1},z_{i-1:i+1})\rmd z_{i-2}\rmd z_{i+1}\eqsp,

@@

where $\psi^n_{i}(a_{i-1:i+1},z_{i-1:i+1})$ is the smoothing distribution of the hidden regimes and states between time indices $i-1$ and $i+1$. Note that the EM algorithm requires the approximation of $p(a_{i-1},z_{i-1},a_{i-1},z_{i-1}|y_{1:n})$ in the E-step, this may be obtained following the same steps by marginalizing explicitly the linear states at time $i-2$ and $i+1$, see Section~\ref{sec:numerical:experiments}. Intermediate computations follow the same steps as  for the approximation of $p(a_{i},z_{i}|y_{1:n})$. $\psi^n_{i}$ may be decomposed as follows:

@@

%\clearpage
%\newpage
%
